% ============================================================
% PROBLEM FORMULATION SECTION
% ============================================================

\section{Problem Formulation}\label{sec:problem}

We formulate sepsis treatment as a finite-horizon Markov Decision Process (MDP) in the offline reinforcement learning setting, where the goal is to learn an optimal policy from a fixed dataset without further environment interaction. We define interpretability through the Linearly Estimated Gradients (LEG) framework for quantitative feature importance measurement.

% ============================================================
\subsection{MDP Formulation}\label{sec:problem:mdp}

The sepsis treatment MDP is defined by the tuple $\mathcal{M} = (\mathcal{S}, \mathcal{A}, \mathcal{P}, \mathcal{R}, \gamma)$. The state space $\mathcal{S} \subset \mathbb{R}^{46}$ captures patient physiological condition through laboratory values, vital signs, and clinical severity scores as detailed in Section~\ref{sec:methods:env}. The action space $\mathcal{A}$ contains 24 discrete actions indexed from 0 to 23, representing a $5 \times 5$ grid of IV fluid and vasopressor dosing levels. The gym-sepsis environment implements the mapping $a = \min(5 \times \text{IV\_bin} + \text{VP\_bin}, 23)$, effectively merging the highest dosage combination into action 23. The transition dynamics $\mathcal{P}(s_{t+1} | s_t, a_t)$ are learned from MIMIC-III data via the gym-sepsis simulator \citep{raghu2017sepsis_drl}. The reward function uses sparse terminal rewards: $\mathcal{R}(s_T, a_T) = +15$ for survival, $-15$ for death, and $0$ for intermediate steps. We use discount factor $\gamma = 0.99$.

A policy $\pi: \mathcal{S} \to \Delta(\mathcal{A})$ maps states to action distributions. The goal is to find $\pi^* = \arg\max_\pi V^\pi(s)$ where the value function is:
\begin{equation}
V^\pi(s) = \mathbb{E}_{\tau \sim \pi, \mathcal{P}} \left[ \sum_{t=0}^{H-1} \gamma^t \mathcal{R}(s_t, a_t) \,\Big|\, s_0 = s \right].
\end{equation}
The action-value function $Q^\pi(s, a)$ represents expected return when taking action $a$ in state $s$ and following $\pi$ thereafter. The optimal policy is derived via $\pi^*(s) = \arg\max_a Q^*(s, a)$.


% ============================================================
\subsection{Offline RL Setting}\label{sec:problem:offline}

In offline RL, the agent learns exclusively from a fixed dataset $\mathcal{D} = \{(s_i, a_i, r_i, s'_i)\}_{i=1}^N$ collected under a behavioral policy, without environment interaction during training \citep{levine2020offline}. The central challenge is distributional shift, where the learned policy $\pi$ may select out-of-distribution actions with unreliable Q-value estimates due to extrapolation error \citep{fujimoto2019offpolicy}.

Conservative Q-Learning \citep{kumar2020cql} addresses this via pessimism by penalizing Q-values for out-of-distribution actions:
\begin{equation}
\min_Q \, \mathbb{E}_{(s,a,r,s') \sim \mathcal{D}} \left[ \left( Q(s, a) - \left( r + \gamma \max_{a'} Q(s', a') \right) \right)^2 \right] + \alpha \cdot \mathbb{E}_{s \sim \mathcal{D}} \left[ \log \sum_{a} \exp(Q(s, a)) - \mathbb{E}_{a \sim \pi_\beta} [Q(s, a)] \right],
\end{equation}
where $\alpha > 0$ controls conservatism strength. Behavior Cloning avoids distributional shift by imitating the behavioral policy via supervised learning: $\pi_{\text{BC}} = \arg\max_\pi \mathbb{E}_{(s, a) \sim \mathcal{D}} [\log \pi(a | s)]$, but cannot improve beyond the behavioral policy's performance.


% ============================================================
\subsection{Interpretability via LEG}\label{sec:problem:interp}

Interpretability quantifies the extent to which clinicians can understand policy decisions. This is critical for regulatory approval \citep{fda2021ai} and clinical trust \citep{holzinger2017xai_healthcare}. We use Linearly Estimated Gradients \citep{greydanus2018leg}, a model-agnostic perturbation method measuring feature importance.

For a policy $\pi$ and state $s$, LEG approximates the saliency of the Q-function with respect to each state feature. We sample $M = 1000$ perturbations $\delta^{(m)} \sim \mathcal{N}(0, \sigma^2 I)$ with $\sigma = 0.05$. We then evaluate perturbed Q-values as $\Delta Q^{(m)} = Q(s + \delta^{(m)}, a) - Q(s, a)$. We fit linear regression $\Delta Q^{(m)} \approx \sum_{j=1}^{46} w_j \cdot \delta_j^{(m)}$ via OLS and extract saliency scores as $\text{Saliency}_j(s, a) = w_j$.

We quantify interpretability using 3 metrics. Maximum saliency magnitude, defined as $\max_{s, j} |\text{Saliency}_j(s, a)|$, measures signal strength. Saliency range captures feature differentiation. Clinical coherence assesses whether top features align with medical knowledge such as blood pressure and lactate for sepsis. Policies with strong saliency signals greater than 10, large ranges, and high clinical coherence are deemed interpretable.

% End of Problem Formulation section