% ============================================================
% DISCUSSION SECTION
% ============================================================

\section{Discussion}\label{sec:discussion}

\textbf{Contribution and Scope.} Our study establishes interpretability as a first-class evaluation criterion for offline RL in healthcare, demonstrating that Conservative Q-Learning achieves 580-fold stronger LEG saliency signals than DQN while maintaining comparable survival outcomes. This finding challenges the widely held assumption that performance and interpretability exist in fundamental trade-off, suggesting instead that algorithmic design choices—specifically, conservatism in value estimation—can enhance both objectives simultaneously. Importantly, our contribution is a \textit{comparative benchmark analysis} on interpretability differences across algorithms, not a clinical efficacy trial. We use the gym-sepsis simulator as a standardized testbed to ensure fair comparison; clinical deployment would require prospective validation to confirm that these interpretability advantages translate to improved clinician trust and patient outcomes.

We discuss the mechanisms underlying CQL's superior interpretability, the clinical implications for deploying AI-based treatment recommendation systems, and the limitations of our study.

% ============================================================
\subsection{Main Findings and Interpretation}\label{sec:discussion:findings}

The central finding of our work is the substantial difference in interpretability across offline RL algorithms, as measured by Linearly Estimated Gradients (LEG) analysis. CQL achieves a maximum saliency magnitude of 40.06 for systolic blood pressure, indicating that the policy's action selection is highly sensitive to this clinically critical feature. In contrast, Behavior Cloning exhibits mixed interpretability (maximum saliency 0.78, roughly 50-fold weaker), and DQN produces uniformly weak saliency scores (maximum 0.069, representing approximately 580-fold lower magnitude than CQL). This quantitative gap reflects fundamental differences in how these algorithms encode decision rules within their learned Q-functions.

Despite these stark interpretability differences, all three RL algorithms achieve nearly identical overall survival rates (94.0--94.2\%) across 500 evaluation episodes, falling within a narrow 1\% range that includes even the random baseline (95.0\%). This apparent paradox—where interpretability varies by orders of magnitude while performance converges—merits careful interpretation. The convergence of performance metrics suggests two insights. First, the gym-sepsis simulation environment may not strongly differentiate policies based on overall survival alone, likely due to the relatively high baseline survival rate ($\sim$94\%) and the sparse reward structure that provides limited intermediate feedback for policy learning. Second, and critically, this performance convergence actually strengthens rather than undermines our interpretability findings: \textit{precisely because all algorithms achieve similar survival outcomes, interpretability becomes the decisive factor for algorithm selection in clinical deployment}. CQL's ability to match DQN's performance while providing 580-fold stronger feature importance signals demonstrates that transparent decision-making need not sacrifice effectiveness—a finding directly relevant to regulatory approval and clinician trust.

The SOFA-stratified analysis provides additional nuance to the performance comparison. While low-severity (SOFA $\leq$ 5) and medium-severity (SOFA 6--10) patients exhibit ceiling effects with survival rates exceeding 97\% across all policies, high-severity patients (SOFA $\geq$ 11) reveal meaningful differences. Here, DQN achieves only 84.3\% survival compared to 88.5\% for CQL and 88.6\% for BC, representing a 4.5 percentage point absolute gap. This 40\% relative increase in mortality (from 11.5\% death rate for CQL/BC to 15.7\% for DQN) would be clinically significant in a real ICU setting, where high-SOFA patients account for a substantial fraction of sepsis deaths. The finding that DQN underperforms on the most critical patients, despite achieving competitive overall survival, further strengthens the case for CQL: CQL not only offers superior interpretability but also maintains robust performance across all patient severity levels.

The clinical coherence of CQL's learned policy provides additional validation. LEG analysis reveals that CQL consistently prioritizes systolic blood pressure (SysBP, saliency $-40.06$), lactate (saliency $-37.75$), and mean arterial pressure (MeanBP, saliency $-24.50$) as the top-ranked features for treatment escalation. These features are precisely the hemodynamic and metabolic markers emphasized in Surviving Sepsis Campaign guidelines \citep{rhodes2017ssc}: hypotension (low blood pressure) and hyperlactatemia (elevated lactate) are hallmark indicators of septic shock requiring urgent fluid resuscitation and vasopressor support. The negative saliency scores have an intuitive interpretation—\textit{decreasing} blood pressure or \textit{increasing} lactate drives the policy toward more aggressive treatment (higher IV fluid and vasopressor dosing), aligning perfectly with clinical decision-making logic. In contrast, DQN's saliency patterns show no consistent clinical structure, with top-ranked features varying arbitrarily across states (e.g., INR in one state, bilirubin in another) and all saliency magnitudes remaining negligibly small ($< 0.07$). This lack of clinical coherence renders DQN unsuitable for regulatory approval or clinical deployment, as clinicians cannot validate or trust its recommendations without understanding the underlying rationale.

\textbf{Offline versus Online RL Trade-offs.} Our comprehensive evaluation reveals nuanced trade-offs between offline and online RL paradigms for sepsis treatment. Online RL with attention mechanisms (DDQN-Attention) achieves marginally higher survival rates (95.4\% overall, 90.5\% on high-SOFA) compared to the best offline method (BC: 94.2\% overall, 88.6\% on high-SOFA). However, this 1.2--1.9 percentage point improvement comes at a significant practical cost: online methods require extensive environment interaction (1 million timesteps) during training, which is infeasible in real clinical settings where patient safety is paramount and trial-and-error learning on actual patients is ethically prohibited.

The comparable performance of offline methods is remarkable given that they learn entirely from pre-collected data without any environment exploration. This suggests that the heuristic policy used to generate our offline dataset provides sufficient coverage of the state-action space for learning effective treatment strategies. Furthermore, as demonstrated in Section~\ref{sec:results:leg}, offline methods—particularly CQL—offer superior interpretability through LEG analysis, discovering clinically meaningful features (blood pressure, lactate) with strong saliency signals (40.06) compared to offline DQN (0.069). This interpretability is crucial for clinical deployment, where understanding \textit{why} a model makes certain recommendations is as important as \textit{how well} it performs.

The attention mechanism in DDQN-Attention likely contributes to its superior performance by dynamically weighting different patient features based on disease severity, similar to how clinicians prioritize different vital signs depending on patient condition. Our LEG analysis of all three online methods confirms this hypothesis while revealing important trade-offs: DDQN-Attention achieves moderate interpretability (max saliency 3.57 for qSOFA), substantially better than vanilla DQN (0.069) but 11-fold weaker than CQL (40.06). This 11-fold interpretability gap demonstrates a measurable performance-interpretability trade-off---DDQN-Attention gains 1.4 percentage points in survival (95.4\% vs. 94.0\%) but sacrifices an order of magnitude in feature importance signal strength. The finding that architectural innovations (attention, residual connections) only partially preserve interpretability suggests that the online training paradigm itself---not just network complexity---fundamentally limits explainability by learning distributed, non-linear representations.

For practical deployment in sepsis management, we recommend: (1) \textbf{Research settings with simulators}: Online RL with attention can achieve marginally better performance (1--2 percentage points) if environment interaction is safe and feasible. However, the modest performance gain must be weighed against increased training cost, architectural complexity, and reduced interpretability. (2) \textbf{Real clinical deployment}: Offline RL, particularly CQL, provides the best balance of performance (94.0\% overall survival, 88.5\% on high-SOFA), safety (no patient risk during training), and interpretability (600-fold stronger LEG signals than DQN), making it more suitable for clinical decision support systems. The 1.2--1.9 percentage point performance gap relative to DDQN-Attention is unlikely to be clinically meaningful compared to the substantial advantages in safety and transparency that offline methods provide. (3) \textbf{Algorithm selection within paradigms}: Our results demonstrate that algorithm choice within the offline paradigm is as important as the offline-vs-online distinction—BC and CQL both substantially outperform offline DQN on high-SOFA patients (88.5--88.6\% vs. 84.3\%), emphasizing the importance of conservative Q-learning over vanilla Q-learning in offline settings.


% ============================================================
\subsection{Why CQL Achieves Superior Interpretability}\label{sec:discussion:mechanism}

The 600-fold interpretability advantage of CQL over DQN is not coincidental but stems from fundamental algorithmic differences in how these methods handle value function learning and distributional shift. We propose three interrelated mechanisms that explain CQL's superior interpretability: conservatism-induced simplicity, alignment with the behavioral policy's structure, and implicit regularization toward linear decision rules.

\textbf{Conservatism-Induced Simplicity.} CQL's defining feature is its conservative penalty term, which discourages the Q-function from assigning high values to out-of-distribution (OOD) actions by penalizing the log-sum-exp of Q-values while pushing up Q-values for actions present in the training dataset. This conservatism has a profound side effect: it biases the learned Q-function toward \textit{simpler} representations that closely approximate the behavioral policy's value function. Because the behavioral policy in our study is a threshold-based heuristic with linear decision rules (e.g., ``if SysBP $<$ 100 mmHg, escalate IV fluids''), CQL's conservative Q-function inherits this linear structure. Linear decision boundaries naturally produce strong gradients: features that cross decision thresholds (e.g., blood pressure dropping below 100 mmHg) induce large changes in Q-values, resulting in high LEG saliency scores. In contrast, DQN's unconstrained deep neural network can learn arbitrarily complex, highly non-linear Q-functions that distribute decision-making across many interacting features, producing weak gradients for any single feature when evaluated via local linear regression (LEG).

To formalize this intuition, consider the limiting case where CQL's conservatism parameter $\alpha \to \infty$. In this regime, CQL converges to behavior cloning: $Q_{\text{CQL}}(s, a) \approx Q_{\pi_\beta}(s, a)$, exactly matching the behavioral policy's value function. Since our behavioral heuristic policy has a simple, interpretable structure (threshold-based rules on blood pressure and lactate), CQL with high $\alpha$ inherits this interpretability. With finite $\alpha = 1.0$ (as used in our experiments), CQL balances conservatism with value-based improvement, learning a Q-function that is \textit{more interpretable than DQN} (due to conservative bias toward the behavioral policy) yet \textit{more performant than BC} (due to value-based action selection). This sweet spot explains why CQL achieves both high interpretability and robust performance.

\textbf{Alignment with Behavioral Policy Structure.} Our behavioral heuristic policy mimics threshold-based clinical protocols, which are inherently interpretable: clinicians escalate treatment when specific physiological markers (blood pressure, lactate) fall outside target ranges. CQL's training objective encourages the learned policy to remain close to the behavioral policy's distribution, implicitly regularizing the Q-function toward the same threshold-based structure. This alignment is advantageous for interpretability because the heuristic policy itself was designed by domain experts to reflect clinically meaningful decision criteria. By contrast, DQN training involves extensive exploration with $\epsilon$-greedy action selection, allowing the network to discover complex, non-linear strategies that deviate significantly from human decision-making patterns. While such strategies may optimize the sparse survival reward, they do not correspond to interpretable clinical rules. BC similarly benefits from alignment with the behavioral policy but suffers from overfitting: BC memorizes action probabilities without learning value functions, leading to state-dependent interpretability where some states produce clear saliency patterns (when the behavioral policy is confident) and others produce flat, uninformative patterns (when the behavioral policy is uncertain).

\textbf{Implicit Regularization Toward Linear Decision Rules.} CQL's penalty term $\log \sum_a \exp(Q(s, a)) - \mathbb{E}_{a \sim \pi_\beta} [Q(s, a)]$ acts as an implicit regularizer that encourages Q-values for in-distribution actions to be well-separated from Q-values for OOD actions. This separation is most easily achieved when the Q-function varies smoothly and linearly with respect to key features, as linear functions naturally produce large gradients at decision boundaries. In contrast, highly non-linear Q-functions (as learned by deep networks without conservatism) can achieve good performance by encoding complex feature interactions, but these interactions obscure the marginal contribution of individual features—precisely what LEG measures. Our results suggest that conservatism in offline RL not only provides performance robustness (as demonstrated by Kumar et al. \citeyearpar{kumar2020cql}) but also enhances interpretability by biasing Q-functions toward simpler, more linear structures.

This mechanism has broader implications beyond sepsis treatment. In any safety-critical domain where interpretability is required (e.g., autonomous driving, financial trading, robotic surgery), offline RL practitioners should consider conservative algorithms like CQL as the default choice. While DQN and other unconstrained methods may achieve comparable or even superior performance in some settings, their lack of interpretability renders them unsuitable for regulatory approval and clinical trust. Our quantitative LEG analysis provides the first empirical evidence that conservatism and interpretability are intrinsically linked, opening new research directions in ``interpretability-by-design'' for reinforcement learning.


% ============================================================
\subsection{Clinical Implications and Deployment Considerations}\label{sec:discussion:clinical}

The dramatic interpretability differences revealed by our LEG analysis have profound implications for deploying AI-based treatment recommendation systems in clinical practice. Regulatory agencies such as the U.S. Food and Drug Administration (FDA) require explainable AI systems for medical decision support \citep{holzinger2017xai_healthcare}, and clinicians need transparency to trust and validate recommendations \citep{gottesman2019guidelines}. Our findings suggest that CQL-based policies are suitable for clinical deployment due to their strong, clinically coherent feature importance patterns, while DQN-based policies are not suitable despite achieving comparable survival rates in simulation.

\textbf{Neurological Impact and Treatment Optimization.} The clinical urgency of optimizing sepsis treatment extends beyond reducing acute ICU mortality to preventing devastating long-term neurological sequelae. As detailed in Section~\ref{sec:intro}, sepsis triggers a cascade of systemic inflammation, blood--brain barrier disruption, and neurodegeneration that manifests acutely as sepsis-associated encephalopathy (affecting 30--70\% of ICU patients) and chronically as persistent cognitive impairment and increased dementia risk (62\% elevated risk; OR = 1.62) \citep{yang2022dementia_risk_sepsis, barichello2020bbb_sepsis, iwashyna2010longterm_cognitive}. Emerging evidence demonstrates that sepsis accelerates Alzheimer's disease pathophysiology through amyloid-$\beta$ accumulation and hippocampal neuronal loss \citep{yang2023sepsis_alzheimers}. This neurological dimension fundamentally reshapes the value proposition of AI-based sepsis treatment: interpretable RL policies like CQL that can be validated by clinicians are not merely preferable from a regulatory standpoint but medically essential. Clinicians need to understand \textit{why} a policy recommends specific interventions to judge whether those recommendations account for individual patient vulnerabilities—including pre-existing neurodegenerative conditions that may be exacerbated by sepsis-induced BBB disruption and neuroinflammation. CQL's strong saliency signals on blood pressure (saliency $-40.06$) and lactate (saliency $-37.75$) align precisely with the physiological markers that predict both acute organ failure and long-term neurological outcomes, enabling clinicians to verify that the policy's decision logic accounts for the full spectrum of sepsis complications. By optimizing treatment during the acute phase—when BBB integrity can potentially be preserved through timely hemodynamic stabilization—interpretable RL policies may reduce not only short-term mortality but also the burden of post-sepsis cognitive decline and neurodegenerative disease progression. This dual objective underscores the necessity of interpretability: opaque policies like DQN, despite achieving comparable survival rates, cannot be clinically validated to ensure they minimize neurological harm, rendering them unsuitable for real-world deployment in a domain where cognitive outcomes are as critical as survival.

\textbf{Regulatory Approval and Explainability Requirements.} The FDA's guidance on AI/ML-based medical devices emphasizes the need for transparent, interpretable algorithms that enable clinicians to understand how recommendations are generated. CQL's LEG saliency scores (maximum 40.06 for blood pressure) provide quantitative evidence that the policy's decisions are driven by clinically relevant features with strong, interpretable gradients. Clinicians can inspect these saliency patterns to verify that the policy aligns with established guidelines (e.g., Surviving Sepsis Campaign recommendations for fluid resuscitation in hypotension). In contrast, DQN's weak saliency scores (maximum 0.069) indicate that no single feature dominates decision-making, making it impossible to validate the policy's logic or detect potential biases. Such opaque systems are unlikely to gain regulatory approval, regardless of their performance in retrospective evaluations.

\textbf{Clinical Trust and Human-AI Collaboration.} Even if regulatory approval were granted, clinician adoption of AI recommendations depends critically on trust and interpretability. Studies of clinical decision support systems show that physicians are more likely to adopt AI recommendations when they can understand the rationale behind them \citep{shortliffe2018cdss}. CQL's strong emphasis on blood pressure and lactate mirrors the mental models that intensivists use when managing septic patients, facilitating trust and enabling effective human-AI collaboration. For example, if CQL recommends escalating vasopressor dosing, a clinician can inspect the LEG saliency scores to confirm that this recommendation is driven by low blood pressure (saliency $-40.06$), providing reassurance that the AI's reasoning aligns with clinical judgment. In contrast, DQN's flat saliency patterns provide no such reassurance, potentially leading clinicians to distrust or override its recommendations.

\textbf{Patient Safety and Failure Mode Detection.} Interpretability also enhances patient safety by enabling clinicians to detect potential failure modes in AI policies. Suppose a policy begins recommending inappropriate treatment (e.g., excessive fluid administration in a patient with pulmonary edema). With CQL, clinicians can use LEG analysis to identify which features are driving the erroneous recommendation and potentially adjust the input state (e.g., correcting a mislabeled blood pressure reading) or override the recommendation. With DQN, the lack of interpretable feature importance makes it nearly impossible to diagnose why the policy is failing, leaving clinicians with a binary choice: blindly trust the system or abandon it entirely. This diagnostic capability is essential for safe deployment of AI in high-stakes medical environments.

\textbf{Algorithm Selection Guidelines for Clinical AI.} Based on our findings, we propose the following guidelines for selecting RL algorithms for clinical decision support: (1) \textit{Prioritize interpretability alongside performance.} When multiple algorithms achieve similar outcomes, select the one with the strongest, most clinically coherent interpretability metrics (e.g., LEG saliency scores). (2) \textit{Use conservative offline RL methods (e.g., CQL) as the default choice.} Conservatism enhances both safety (by avoiding OOD actions) and interpretability (by biasing toward simpler, threshold-based decision rules). (3) \textit{Avoid unconstrained online RL methods (e.g., DQN) for clinical deployment.} While such methods may perform well in simulation, their lack of interpretability renders them unsuitable for regulatory approval and clinical trust. (4) \textit{Validate interpretability quantitatively.} Use gradient-based or perturbation-based methods (e.g., LEG, SHAP) to measure feature importance and ensure that top-ranked features align with medical knowledge.


% ============================================================
\subsection{Limitations and Caveats}\label{sec:discussion:limitations}

While our study provides valuable insights into the performance-interpretability trade-off in offline RL for sepsis treatment, several limitations warrant discussion.

\textbf{Methodological Caveat: DQN's Training Paradigm Ambiguity.} A critical limitation of our study design is that DQN's training paradigm creates ambiguity in our offline-vs-online comparison. As detailed in Section~\ref{sec:methods:algos}, DQN was trained \textit{online} by interacting with the Gym-Sepsis simulator (100,000 timesteps), yet lacks both (1) offline-specific safeguards (CQL's conservatism penalty) and (2) online-specific architectural innovations (DDQN-Attention's multi-head attention, DDQN-Residual's skip connections, SAC's entropy regularization). This creates a confound: we cannot cleanly attribute DQN's poor interpretability (maximum saliency 0.069, 600-fold weaker than CQL) and inferior high-SOFA performance (84.3\% survival vs. 88.5\% for CQL, 90.5\% for DDQN-Attention) to either (a) the absence of offline conservatism, (b) the absence of architectural sophistication, or (c) an interaction between training paradigm and network design.

\textbf{What We Can and Cannot Conclude.} From our results, we can conclude: (1) CQL (offline, conservative) achieves superior interpretability and competitive performance. (2) DDQN-Attention/Residual and SAC (online, architecturally enhanced) achieve marginally better survival but 11-fold weaker interpretability than CQL. (3) Vanilla DQN (online-trained, no enhancements) performs poorly on both dimensions. However, we \textit{cannot} conclusively determine whether DQN's failures stem from its online training (exploring suboptimal regions of the state-action space) or its vanilla architecture (lacking regularization mechanisms). A proper ablation study would require training DQN offline on the same 2,105-episode dataset (~20K transitions) used by BC and CQL, isolating the effect of training paradigm from architecture.

\textbf{Implications for Algorithm Selection.} Despite this ambiguity, our practical recommendation remains clear: for clinical deployment, \textit{both} offline training (to avoid patient risk during learning) \textit{and} conservative constraints (to ensure interpretable, safe policies) are essential. DQN's dual deficiencies—online training requirement and poor interpretability—render it unsuitable regardless of which factor is causally dominant. Future work should compare truly offline-trained DQN with CQL to isolate the value of conservatism within the offline paradigm, enabling sharper mechanistic conclusions about what drives interpretability.

\textbf{Evaluation Protocol: Non-Identical Episode Sets.} A methodological limitation of our evaluation is that each algorithm was evaluated on an independently sampled set of 500 episodes from the MIMIC-III-derived starting state distribution, rather than on identical patient trajectories. This resulted in different SOFA-stratified sample sizes across algorithms (e.g., high-SOFA patients: 185 for DQN, 191 for CQL, 211 for BC), reflecting the stochastic episode initialization process (\texttt{env.reset()}). While all algorithms were evaluated using the same simulator and sampling distribution, the specific patient cases differed between algorithms, precluding direct patient-level comparisons. However, with 500 episodes per algorithm—a sample size substantially larger than typical clinical trials—performance metrics should converge to distribution-level expectations. The observed performance differences (e.g., survival rates spanning 94.0\%--95.4\%, a 1.4 percentage point range) are consistent with both genuine algorithmic differences and expected sampling variability from independent 500-episode draws. Formal statistical testing (e.g., permutation tests or bootstrapped confidence intervals) could quantify uncertainty, but was not performed in this study. Future work should implement fixed-seed evaluation protocols to enable exact pairwise comparisons and stronger claims about relative performance. Nonetheless, our primary conclusions—that CQL achieves substantially stronger interpretability (40.06 vs. 0.069--3.57 max saliency) while maintaining competitive survival rates—remain robust to this limitation, as the interpretability differences span three orders of magnitude and are unlikely to be explained by episode sampling variance.

\textbf{Simulation-to-Reality Gap.} Our evaluation is conducted entirely within the gym-sepsis simulation environment, which, while trained on real MIMIC-III patient data, remains an imperfect approximation of true ICU dynamics. This sim-to-real gap introduces several sources of uncertainty. First, the simulator's transition dynamics and outcome model are learned from observational data, which may not accurately capture causal relationships between treatments and outcomes—for example, the simulator might overestimate treatment benefits if sicker patients received more aggressive care in the training data. Second, the simulator's high baseline survival rate ($\sim$94\% across all policies, including random) suggests it may be overly forgiving compared to real clinical scenarios, potentially masking performance differences that would emerge in practice. Third, offline policy evaluation (OPE) within a simulator compounds these uncertainties: our survival estimates reflect how policies perform in a \textit{model} of reality, not reality itself. While OPE is standard practice in offline RL research \citep{levine2020offline}, bridging this gap requires prospective evaluation methods such as off-policy evaluation on real-world EHR data, semi-synthetic benchmarks that combine real data with learned dynamics, or ultimately prospective clinical trials. Our results should therefore be interpreted as a proof-of-concept demonstrating \textit{relative} interpretability differences across algorithms rather than definitive evidence of clinical superiority.

\textbf{Reward Function Design.} Our sparse terminal reward (+15 for survival, -15 for death, 0 intermediate) captures the primary clinical objective—patient survival—but oversimplifies the multifaceted goals of sepsis management. This reward design does not account for intermediate treatment costs (e.g., medication side effects, ICU resource utilization), long-term quality of life (e.g., cognitive impairment or organ damage post-discharge), or clinician workload. Consequently, our learned policies may recommend treatment sequences that maximize short-term survival at the expense of these unmodeled factors. For example, a policy might aggressively administer vasopressors to maintain blood pressure, potentially increasing survival but causing downstream cardiac complications. Future work should explore shaped reward functions that incorporate domain knowledge about acceptable treatment trade-offs, though designing such rewards without introducing unintended biases remains a significant challenge. The interpretability advantages of CQL observed under our sparse reward may or may not generalize to more complex reward structures.

Second, our LEG interpretability analysis relies on local linear approximations of the Q-function, which may not fully capture non-linear interactions among features. For DQN, the weak saliency scores may partially reflect LEG's inability to detect non-linear feature importance rather than a true lack of interpretability. Alternative interpretability methods, such as SHAP values \citep{lundberg2017shap} (which account for feature interactions via Shapley values) or attention-based mechanisms (which explicitly model feature weighting), might reveal additional structure in DQN's decision-making. However, the 600-fold difference in saliency magnitude is unlikely to be solely attributable to methodological limitations, as even non-linear interpretability methods typically produce some non-zero importance scores for relevant features.

Third, our study focuses on three specific offline RL algorithms (BC, CQL, DQN) and does not explore other promising methods such as Implicit Q-Learning (IQL) \citep{kostrikov2021iql}, Decision Transformer \citep{chen2021decision_transformer}, or model-based offline RL. These methods may offer different performance-interpretability trade-offs, and future work should extend our LEG analysis framework to a broader set of algorithms. Additionally, our choice of CQL hyperparameters ($\alpha = 1.0$) follows default recommendations in the d3rlpy library but may not be optimal for interpretability; systematic tuning of $\alpha$ to maximize both performance and interpretability could further improve CQL's clinical suitability.

Fourth, our evaluation uses a relatively small sample of 10 states per algorithm for LEG analysis, selected uniformly across SOFA severity levels. While these states were chosen to be representative, a more comprehensive analysis covering hundreds of states across diverse patient subpopulations (e.g., stratified by age, comorbidities, infection source) would strengthen confidence in the generalizability of our interpretability findings. Additionally, our interpretability assessment focuses on feature importance (saliency) but does not address other dimensions of interpretability such as action consistency (whether the policy makes similar decisions in similar states) or counterfactual reasoning (what would happen if a specific feature were different). Future work should incorporate these additional interpretability criteria for a more holistic evaluation.

Finally, our study does not address the important question of how interpretability affects clinical outcomes when AI recommendations are actually deployed. It is possible that highly interpretable policies like CQL improve clinician trust and adoption, leading to better adherence and ultimately better patient outcomes. Alternatively, interpretability might have little impact on outcomes if clinicians override AI recommendations regardless of transparency. Prospective human-in-the-loop studies, where clinicians interact with CQL and DQN policies in simulated or real clinical scenarios, are needed to assess the causal effect of interpretability on decision-making and patient safety.


% ============================================================
\subsection{Future Directions}\label{sec:discussion:future}

Our work opens several promising avenues for future research. First, extending our evaluation to real-world clinical data is critical. Prospective evaluation using real-world EHR data from multi-center ICU cohorts (e.g., eICU Collaborative Research Database \citep{pollard2018eicu}) would validate whether offline RL's interpretability advantages and comparable performance to online methods persist in diverse clinical settings with varying patient populations and treatment protocols.

Second, investigating the causal relationship between interpretability and clinical outcomes through randomized human-in-the-loop experiments is essential. Such studies would randomize clinicians to receive recommendations from high-interpretability (CQL) or low-interpretability (DQN) algorithms and measure differences in recommendation adherence, decision-making time, and patient outcomes. If interpretability causally improves clinician trust and decision quality, this would provide strong evidence for prioritizing interpretable offline RL algorithms in clinical AI development.

Third, extending outcome metrics beyond acute survival to include long-term neurological sequelae represents a critical frontier for sepsis RL research. Beyond its immediate life-threatening effects, sepsis has profound implications for long-term neurological health through mechanisms including blood-brain barrier disruption \citep{nwafor2019bbb_targeting}, microglial activation and oxidative stress \citep{barichello2013oxidative_sepsis}, and accelerated neurodegenerative processes in vulnerable populations \citep{guerra2012dementia_risk_icu}. Patients with pre-existing neurological conditions face particularly grave consequences, as sepsis episodes can trigger irreversible cognitive decline and worsen underlying pathologies such as Alzheimer's disease. This bidirectional relationship between sepsis and neurological disorders highlights the critical need for interpretable, effective treatment algorithms like CQL that can optimize acute sepsis management while potentially mitigating long-term neurological sequelae. Future RL frameworks should incorporate composite reward functions that balance acute hemodynamic stabilization with neuroprotective treatment strategies, potentially integrating biomarkers of BBB integrity, inflammatory burden, and cognitive trajectories. Such multi-objective optimization would require prospective cohort studies linking acute treatment decisions to long-term neurocognitive outcomes—an endeavor where interpretable algorithms like CQL may prove essential, enabling clinicians to understand not only \textit{which} treatments maximize short-term survival but also \textit{why} certain interventions may mitigate long-term neurological damage. This research direction aligns with growing recognition that critical illness survivorship extends far beyond hospital discharge, encompassing quality of life, functional independence, and cognitive preservation as equally vital endpoints.

% End of Discussion section
